Exercise Manner Prediction based on HAR Dataset
========================================================

This analysis builds a model for predicting Exercise Manner using HAR data from http://groupware.les.inf.puc-rio.br/har for the purpose of Coursera's Practical Machine Learning class.

Initialization
--------------
The training of the model is performed using `caret` package. 
In order to speed-up training phase `doMC` package is also loaded.

```{r cache=T}
library(caret)
library(doMC)
registerDoMC(cores = 4)
```

The seed is set to an arbitrary value.
```{r cache=T}
set.seed(12345)
```

Data preprocessing
------------------
The data is obtained with the following commands.

```{r, cache=TRUE}
download.file('https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv', destfile='pml-training.csv', method="curl")
download.file('https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv', destfile='pml-testing.csv', method="curl")
```
```{r, cache=TRUE}
pml.training <- read.csv('pml-training.csv', header=T)
pml.testing <- read.csv('pml-testing.csv', header=T)
```

A clean dataset is prepared by filtering out variables that can be derived from the other ones (e.g. `kurtosis`, `skewness`) and the `X` variable that enumerates samples.

```{r, cache=TRUE}
skipped.cols <- '^stddev_|^var_|^kurtosis_|^skewness_|^amplitude_|^min_|^max_|^avg_|^X$|^cvtd_'

cleaned.training <- pml.training[, -grep(skipped.cols, names(pml.training))]
cleaned.testing<- pml.testing[, -grep(skipped.cols, names(pml.testing))]
```

The `cleaned.training` dataset is split into two sets - `training` and `testing`. 

```{r, cache=TRUE}
trainingIndex <- createDataPartition(cleaned.training$classe, p=0.75, list=F)
training <- cleaned.training[trainingIndex,]
testing <- cleaned.training[-trainingIndex,]
```

The `training` set will be used to train the model.  The `testing` set will be used to estimate out-of-sample error.

Training and Parameter Estimation
---------------------------------
For the purpose of prediction the random forest model has been chosen since it is a state-of-art statistical learning method.

First, we perform *5*-fold cross-validation training to estimate the `mtry` parameter of the random forest model.
```{r cache=T}
fitControl <- trainControl(method = "repeatedcv", number = 5, repeats = 5, verboseIter = T)
dev.model <- train(classe ~ ., method="rf", data=training, trControl=fitControl)
```

Having the model trained we estimate out-of-sample error by predicting the values using the held-out data from our `testing` set and measuring accuracy. 

```{r cache=T}
testing.predicted <- predict(dev.model, testing)
```
As can be read from the confusion matrix below, accuracy is satisfactory on the testing set.

```{r cache=T}
cm <- confusionMatrix(testing.predicted, testing$classe)
cm
```

The out-of-sample error rate is as follows.
```{r cache=T}
1 - cm$overall[["Accuracy"]]
```

Prediction
-----------
For the purpose of predicting labels for the test cases from the `pml.testing` dataset the final model is build using all the available data by performing the same cross-validation procedure as above.

```{r cache=T}
final.model <- train(classe ~ ., method="rf", data=cleaned.training, trControl=fitControl)
```

The model predicts the following classes for the `pml.testing` dataset.

```{r cache=T}
pml.testing.predicted <- predict(final.model, cleaned.testing)
pml.testing.predicted
```

Finally, the submission files are generated using the procedure `pml_write_files` taken from the PML course site.

```{r cache=T}
pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}

pml_write_files(pml.testing.predicted)
```
